The Sub-Millimetre User Reduction Facility: an iterative map-maker for SCUBA-2

### 2012-06-06 smurf_paper.tex
re-made point source maps, update text for change to 100 from 50 iterations

The first, simplest reduction of the data uses only the \model{COM}
model to estimate and remove the common-mode signal in order to
suppress low-frequency noise in the data. After \model{COM}, the
extinction correction is applied (\model{EXT}), and an initial map is
estimated using equal weighting for all of the detectors. This
estimate of \model{AST} is then removed from the data, and the noise
is measured in the residuals to estimate weights for the subsequent
and final iteration. The resulting map after these two iterations is
shown in Fig.~\ref{fig:pointmaps}a. While the S/N of Uranus is clearly
large ($\sim$175 as estimated by SMURF), enabling us to see the faint
sidelobes (circle and cross pattern), the map also has obvious
circular streaks and other large-scale ripples. The circular streaks
are due to the fact that \model{COM} does not account for all of the
low-frequency noise (see Fig.~\ref{fig:pspec}), and therefore each
bolometer leaves a trace of the circular scan pattern in the map as
each of their baselines slowly drift independently. Most of the
larger-scale ripples in the map, however, are probably due to
degeneracies in the map solution as we discuss below.

To illustrate how large-scale ripples can form (and grow), the same
map solution is run for 50 iterations and shown in
Fig.~\ref{fig:pointmaps}b, now exhibiting a strong vertical
gradient. The degeneracy is easy to understand if the time-domain
behaviour of each model component is considered. The top panel of
Fig.~\ref{fig:degeneracy} shows the residual signals for a single
bolometer after 2 (black) and 50 (grey) iterations which are nearly
identical, yet the \emph{change} in the estimated \model{COM} (green)
and \model{AST} (red) signals between 2 and 50 iterations are
large. However, it is also clear that the estimated \model{COM} and
\model{AST} signals are \emph{complimentary}. In other words, the
large change in the \model{AST} signal is cancelled by freedom in the
\model{COM} signal to grow with opposite sign. For comparison, the
bottom panel of Fig.~\ref{fig:degeneracy} shows the pointing solution
for this section of data, and the shapes of the \model{AST} and
\model{COM} signals match the elevation component, which is aligned
with the gradient in Fig.~\ref{fig:pointmaps}b. Generically, the
calculation of \model{COM} will remove any information on angular
scales that are larger than the array footprint (outline shown in
Fig.~\ref{fig:pointmaps}b for reference), meaning that the map
solution is unconstrained on such large scales.

One improvement is to simply restrict the low-frequency signal that
goes into the maps by applying a high-pass filter to the bolometer
time-series. A third reduction of the data uses the default map-making
parameters, as described in Section~\ref{sec:algorithm}, which adds
the \model{FLT} model to accomplish this task immediately prior to map
estimation. We set the filter edge based on an angular scale of
200\,arcsec, which given the scan speed of 155\,arcsec\,sec$^{-1}$,
corresponds to a frequency of 0.78\,Hz. Now using the automatic
map-based convergence test (Section~\ref{sec:converge}), the resulting
map converges after 10 iterations (Fig.~\ref{fig:pointmaps}c). Both
the circular streaks and the large-scale gradient are now removed, but
they have been replaced by an obvious, circularly-symmetric ringing
pattern about Uranus. The reason for this ringing is that the
hard-edged high-pass filter in frequency space is equivalent to a
$\sinc$ function-like response in map space. Since the scan pattern is
isotropic (scans equally at all position angles), and there is a
bright point-like source, the result is an azimuthally-symmetric
$\sinc$ function-like pattern in the map about Uranus. Note that due
to the iterative nature of the solution, much of the signal in the
source is actually removed for subsequent iterations of \model{FLT},
but as in the previous case, degeneracies between the map and the
\model{FLT} and \model{COM} signal components can still lead to
complicated structures.

The preceding examples illustrate the need for constraints in the map
solution in many situations. For calibrators (and other previously
known bright, compact sources), a good, simple prior is to constrain
the map to a value of zero away from the known locations of
emission. In Fig.~\ref{fig:pointmaps}d, a map is produced in an
identical manner to the previous example, but now setting the map
explicitly to zero beyond a radius of 60\,arcsec from the location of
Uranus (much larger than the FWHM of the main lobe), for all but the
final iteration. In this case, the map converges after 22 iterations,
and all of the previous artifacts have been effectively removed. Since
the map is now flat away from the source, and constrained to a value
of zero, it is appropriate for performing aperture photometry
directly, with no need for an additional reference annulus. This
approach to constrained map-making is similar to that employed for
poorly cross-linked scans of compact (though resolved) sources by
\citet{wiebe2009} using BLAST data.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pointmaps.pdf}
\caption{Multiple reductions of a 450\,\micron\ daisy scan of Uranus,
  all scaled between -0.002\,pW (white) and +0.002\,pW (black): (a)
  Reduction in which only common-mode subtraction is used to suppress
  low-frequency noise, and the reduction is forced to use 2 iterations
  (after the first iteration an estimate of the source flux is
  removed, and the noise is measured in the residuals to obtain
  appropriate weighting for the second and final iteration) ---
  circular streaks are caused by independent low-frequency noise that
  is not removed by the common-mode (Uranus peak 0.14\,pW); (b) same
  as (a) but using 50 iterations, illustrating the degeneracy between
  large-scale structure and the common-mode (the footprint of working
  bolometers is also shown for reference, Uranus peak 0.14\,pW); (c)
  reduction in which high-pass filtering above 0.775\,Hz
  (corresponding to a spatial scale of 200\,arcsec as indicated) is
  applied after common-mode removal, but before the map estimate ---
  the map-based convergence test is activated and the solution halts
  after 10 iterations, but leaving large-scale ringing due to the
  filter (Uranus peak 0.38\,pW); (d) same as (c), but now the region
  of the map beyond the white dot-dashed circle is constrained to a
  value of zero until all but the final iteration --- the map is now
  extremely flat, there is little attenuation of the source flux, and
  the diffraction pattern is clearly seen (Uranus peak 0.38\,pW).}
\label{fig:pointmaps}
\end{figure*}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{degeneracy.pdf}
\caption{Demonstration of the degeneracy between large-scale
  structures in the map and common-mode removal, corresponding to
  panels (a) and (b) in Fig.~\ref{fig:pointmaps}. The black and grey
  lines in the top panel show the residual signal for a single
  bolometer after 2 and 50 iterations respectively (they are nearly
  identical). The green and red lines show the \emph{difference}
  between the \model{COM} and \model{AST} (the signal produced by the
  current map estimate for a given bolometer) model components for
  that bolometer between iterations 2 and 50 respectively. This shows
  that a strong signal has grown over time, and it has equal but
  opposite signs in the two model components, so that they cancel one
  another. The bottom panel shows the scan pattern of the telescope
  over the same period; clearly the \model{COM}/\model{AST} model
  degeneracy is correlated with the elevation offset, and referring to
  Fig.~\ref{fig:pointmaps}, panel (b), this clearly corresponds to the
  strong vertical gradient that has appeared}.
\label{fig:degeneracy}
\end{figure}

>>>>

The first, simplest reduction of the data uses only the \model{COM}
model to estimate and remove the common-mode signal in order to
suppress low-frequency noise in the data. After \model{COM}, the
extinction correction is applied (\model{EXT}), and an initial map is
estimated using equal weighting for all of the detectors. This
estimate of \model{AST} is then removed from the data, and the noise
is measured in the residuals to estimate weights for the subsequent
and final iteration. The resulting map after these two iterations is
shown in Fig.~\ref{fig:pointmaps}a. While the S/N of Uranus is clearly
large ($\sim$250 as estimated by SMURF), enabling us to see the faint
sidelobes (circle and cross pattern), the map also has obvious
circular streaks and other large-scale ripples. The circular streaks
are due to the fact that \model{COM} does not account for all of the
low-frequency noise (see Fig.~\ref{fig:pspec}), and therefore each
bolometer leaves a trace of the circular scan pattern in the map as
each of their baselines slowly drift independently. A significant
contribution to the larger-scale ripples in the map, however, can be
made by degeneracies in the map solution as we discuss below.

To illustrate how large-scale ripples can form (and grow), the same
map solution is run for 100 iterations and shown in
Fig.~\ref{fig:pointmaps}b, now exhibiting a strong vertical
gradient. The degeneracy is easy to understand if the time-domain
behaviour of each model component is considered. The top panel of
Fig.~\ref{fig:degeneracy} shows the residual signals for a single
bolometer after 2 (black) and 100 (grey) iterations which are nearly
identical, yet the \emph{change} in the estimated \model{COM} (green)
and \model{AST} (red) signals between 2 and 100 iterations are
large. However, it is also clear that the estimated \model{COM} and
\model{AST} signals are \emph{complimentary}. In other words, the
large change in the \model{AST} signal is cancelled by freedom in the
\model{COM} signal to grow with opposite sign. For comparison, the
bottom panel of Fig.~\ref{fig:degeneracy} shows the pointing solution
for this section of data, and the shapes of the \model{AST} and
\model{COM} signals match the elevation component, which is aligned
with the gradient in Fig.~\ref{fig:pointmaps}b. Generically, the
calculation of \model{COM} will remove any information on angular
scales that are larger than the array footprint (outline shown in
Fig.~\ref{fig:pointmaps}b for reference), meaning that the map
solution is unconstrained on such large scales.

One improvement is to simply restrict the low-frequency signal that
goes into the maps by applying a high-pass filter to the bolometer
time-series. A third reduction of the data uses the default map-making
parameters, as described in Section~\ref{sec:algorithm}, which adds
the \model{FLT} model to accomplish this task immediately prior to map
estimation. We set the filter edge based on an angular scale of
200\,arcsec, which, given the scan speed of 155\,arcsec\,sec$^{-1}$,
corresponds to a frequency of 0.78\,Hz. Now using the automatic
map-based convergence test (Section~\ref{sec:converge}), the resulting
map converges after 10 iterations (Fig.~\ref{fig:pointmaps}c). Both
the circular streaks and the large-scale gradient are now removed, but
they have been replaced by an obvious, circularly-symmetric ringing
pattern about Uranus. The reason for this ringing is that the
hard-edged high-pass filter in frequency space is equivalent to a
$\sinc$ function-like response in map space. Since the scan pattern is
isotropic (scans equally at all position angles), and there is a
bright point-like source, the result is an azimuthally-symmetric
$\sinc$ function-like pattern in the map. Note that due to the
iterative nature of the solution, much of the signal in the source is
actually removed for subsequent iterations of \model{FLT}, but as in
the previous case, degeneracies between the map and the \model{FLT}
and \model{COM} signal components can still lead to complicated
structures.

The preceding examples illustrate the need for constraints in the map
solution in many situations. For calibrators (and other previously
known bright, compact sources), a good, simple prior is to constrain
the map to a value of zero away from the known locations of
emission. In Fig.~\ref{fig:pointmaps}d, a map is produced in an
identical manner to the previous example, but now setting the map
explicitly to zero beyond a radius of 60\,arcsec from the location of
Uranus (much larger than the FWHM of the main lobe), for all but the
final iteration. In this case, the map converges after 6 iterations,
and all of the previous artifacts have been effectively removed. Since
the map is now flat away from the source, and constrained to a value
of zero, it is appropriate for performing aperture photometry
directly, with no need for an additional reference annulus. This
approach to constrained map-making is similar to that employed for
poorly cross-linked scans of compact (though resolved) sources by
\citet{wiebe2009} using BLAST data.

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pointmaps.pdf}
\caption{Multiple reductions of a 450\,\micron\ daisy scan of Uranus,
  all scaled between -0.002\,pW (white) and +0.002\,pW (black): (a)
  Reduction in which only common-mode subtraction is used to suppress
  low-frequency noise, and the reduction is forced to use 2 iterations
  (after the first iteration an estimate of the source flux is
  removed, and the noise is measured in the residuals to obtain
  appropriate weighting for the second and final iteration) ---
  circular streaks are caused by independent low-frequency noise that
  is not removed by the common-mode (Uranus peak 0.15\,pW); (b) same
  as (a) but using 50 iterations, illustrating the degeneracy between
  large-scale structure and the common-mode (the footprint of working
  bolometers is also shown for reference, Uranus peak 0.15\,pW); (c)
  reduction in which high-pass filtering above 0.775\,Hz
  (corresponding to a spatial scale of 200\,arcsec as indicated) is
  applied after common-mode removal, but before the map estimate ---
  the map-based convergence test is activated and the solution halts
  after 10 iterations, but leaving large-scale ringing due to the
  filter (Uranus peak 0.27\,pW); (d) same as (c), but now the region
  of the map beyond the white dot-dashed circle is constrained to a
  value of zero until all but the final iteration --- the map is now
  extremely flat, there is little attenuation of the source flux, and
  the diffraction pattern is clearly seen (Uranus peak 0.27\,pW).}
\label{fig:pointmaps}
\end{figure*}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{degeneracy.pdf}
\caption{Demonstration of the degeneracy between large-scale
  structures in the map and common-mode removal, corresponding to
  panels (a) and (b) in Fig.~\ref{fig:pointmaps}. The black and grey
  lines in the top panel show the residual signal for a single
  bolometer after 2 and 100 iterations respectively (they are nearly
  identical). The green and red lines show the \emph{difference}
  between the \model{COM} and \model{AST} (the signal produced by the
  current map estimate for a given bolometer) model components for
  that bolometer between iterations 2 and 100 respectively. This shows
  that a strong signal has grown over time, and it has equal but
  opposite signs in the two model components, so that they cancel one
  another. The bottom panel shows the scan pattern of the telescope
  over the same period; clearly the \model{COM}/\model{AST} model
  degeneracy is correlated with the elevation offset, and referring to
  Fig.~\ref{fig:pointmaps}, panel (b), this clearly corresponds to the
  strong vertical gradient that has appeared}.
\label{fig:degeneracy}
\end{figure}
